---
title: Submitting a Post with Blogdown
author: Thomas Roh
date: '2017-09-11'
slug: submitting-a-post-with-blogdown
categories:
  - R
  - tutorial
tags:
  - blogdown
  - R Markdown
  - github
  - git
  - RStudio
---

## Set up

You will first need to set up an account on [**github**](https://www.github.com) and 
install [**git**](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git).

Next, make a copy of the repository into your desired workspace. I like to keep
all of my version controlled projects under one directory. Open up a command
terminal of your choosing and use the following:

```{bash, eval = FALSE}
cd [working directory]
git clone https://github.com/omahar/omahar.git
git branch [new branch]
git checkout [new branch]
git push --set-upstream origin development
```


The [`blogdown`](https://bookdown.org/yihui/blogdown/) package is an R interface 
to creating static web pages with Hugo. I will be using the 
[**RStudio**](https://www.rstudio.com/) IDE and 
this package to do the majority of the heavy lifting.

```{r, eval = FALSE}
devtools::install_github('rstudio/blogdown')
```


## Creating a New Post

## Submitting a Post to the Website Repository

## Workflow

```{r}
library(DiagrammeR)
dotText <- "

digraph workflow {

  graph[rankdir = LR]

  'git pull' -> 'new_post'
  'new_post' -> 'git commit'
  'git commit' -> 'git push'
  'git push' -> 'pull request'
  'pull request' -> 'review'
  'review' -> 'git commit'
  'review' -> 'merge'
  'merge' -> 'netlify'
  'netlify' -> 'new post'
  

}
"
grViz(dotText)

```


