---
title: NCAA College Bowl Pick'em Predictions
author: Thomas Roh
date: '2018-01-24'
slug: ncaa-college-bowl-pick-em-predictions
categories: []
tags: 
  -R
  -gbm
  -rvest
  -dplyr
  -tidyr
  -stringr
  -caret
  -stepAIC
---

***Please read the [Terms of Use](http://www.cfbstats.com/blog/terms-of-use/) if you are going to use the data**

This is the code from my presentation last week at the Omaha R User Group Meeting. 
A fun competition every year is to pick the winners for each college football 
bowl. I wanted to build a data-driven model to help me make this year's picks. 
This post will walk through the process of obtaining data for predictions, 
building a couple predictive models, and then compare the results of the 
models.

### Scraping Data

I'm using the [`rvest`](https://cran.r-project.org/web/packages/rvest/index.html) 
to scrape some data that has been made available to the public at www.cfbstats.com. 
I recommend looking up some examples and using [SelectorGadget](http://selectorgadget.com/) extension to look up css or xpath 
for queries.

First, I want to get a list of all the NCAA FBS teams.

```{r, message=FALSE}
library(rvest)
library(stringr)
library(dplyr)
library(tidyr)
library(caret)

cfbStats <- read_html("http://www.cfbstats.com/")
teams <- cfbStats %>%
  html_nodes(".sub1 a") %>%
  html_text()
teams[1:5]
```

You can click on a team and see some of the underlying data that is available. 
There are 130 teams so going through each link and acquiring the data is fairly 
time expensive. Instead, I'm going to acquire a vector of all of those links.


```{r}
cfbLinks <- cfbStats %>%
  html_nodes(".sub1 a") %>%
  html_attr("href") %>%
  paste0("http://www.cfbstats.com", .)
cfbLinks[1:5]
```

All of the data points are contained with html tables so I'm going to loop through 
and grab the tables for each team.

```{r}
cfbTables <- lapply(cfbLinks, function(site) {
  Sys.sleep(1)
  read_html(site) %>%
    html_table()
})
```

### Data Cleaning

I looked at the statistics provided for each of the tables and I came up with a
list of the ones that I wanted.

```{r}
statNames <- c("Points_per_Game","Total_First_Downs","Rushing_Yards_Mean",
                "Passing_Rating","Passing_Att_Comp_INT_TD","Total_Offense_YDs_per_Play",
                "Total_Offense_YDs","Punt_Returns","Kickoff_Returns","Punting_Yards",
                "INT_INTYards_INTTD","Fumbles_Total_Lost","Penalties_Total_Yards",
                "Third_Down","Fourth_Down","Red_Zone","Field_Goals","PAT_Kicking")

# overall season team stats
whichStats <- c(1,3,5,7,9,10,11,12,14,16,18,19,20,22,24,26,28,30)
teamStats <- lapply(cfbTables, function(listElement) {
  setNames(listElement[[1]][whichStats, ], c("Stat","Home","Opp"))
}) %>%
  setNames(teams) %>%
  bind_rows(.id = "teamName") %>%
  mutate(Stat = rep(statNames, length(teams)))
```

The data isn't structured the way I need it yet and numbers are in an awkward format.

```{r}
home <- teamStats %>%
  select(-Opp) %>%
  gather(team, value,-teamName,-Stat) %>%
  spread(Stat, value) %>%
  transmute(
    teamName,
    team,
    FG_Perc = str_extract(Field_Goals, "[0-9]+"),
    Fourth_Down_Perc = str_extract(Fourth_Down, "[0-9]+"),
    Fumbles_Lost = str_extract(Fumbles_Total_Lost, "[0-9]+$"),
    Passing_Rating = Passing_Rating,
    INT = str_extract(INT_INTYards_INTTD, "^[0-9]+"),
    Avg_Kickoff_Returns = Kickoff_Returns,
    PAT_Perc = str_extract(PAT_Kicking, "[0-9]+"),
    Penalty_Yards = str_extract(Penalties_Total_Yards, "[0-9]+$"),
    Points_per_Game,
    Avg_Punt_Returns = Punt_Returns,
    Avg_Punting = Punting_Yards,
    Red_Zone_Perc = str_extract(Red_Zone, "[0-9]+"),
    Avg_Rush_Yards = Rushing_Yards_Mean,
    Third_Down_Perc = str_extract(Third_Down, "[0-9]+"),
    Total_First_Downs,
    Total_Offense_YDs_per_Play,
    Total_Offense_YDs = str_extract(Total_Offense_YDs, "[0-9]+$")
  ) %>%
  gather(stat, value, -teamName, -team) %>%
  mutate(stat = tolower(paste(team, stat, sep = "_"))) %>%
  select(-team) %>%
  spread(stat, value)
home[ ,-1] <- sapply(home[ ,-1], as.numeric)

opp <- setNames(home, sub("home_", "opp_", names(home)))
```

Now, I have all the team stats that I selected and I can easily join the stats to
the regular season games. The 2nd html table for each team contains their regular 
season schedule.

```{r}
regSeason <- lapply(cfbTables, `[[`, 2) %>%
  setNames(teams) %>% 
  bind_rows(.id = "teamName") %>%
  filter(!is.na(Result),!str_detect(Opponent,"\\+")) %>%
  mutate(opponent = str_replace(Opponent,"@ ",""), 
         opponent = str_replace(opponent,"[0-9]+ ",""),
         win = ifelse(str_detect(Result,"W"),1,0),
         score = str_extract(Result,"[0-9]+") %>% as.numeric(),
         opponent_score = as.numeric(str_extract(Result,"[0-9]+$"))) %>%
  dplyr::select(teamName,opponent,win,score,opponent_score)

regSeason$opponent <- str_replace_all(
  regSeason$opponent,
  c(
  "St\\." = "State",
  "Ga\\." = "Georgia",
  "La\\." = "Louisiana",
  "Caro\\." = "Carolina",
  "Fla\\." = "Florida",
  "Ill\\." = "Illinois",
  "Mich\\." = "Michigan",
  "Miss\\." = "Mississippi",
  "Ala\\." = "Alabama",
  "Ky\\." = "Kentucky",
  "Tenn\\." = "Tennessee",
  "Hawaii" = "Hawai'i",
  "Brigham Young" = "BYU",
  "Southern California" = "USC",
  "Southern Methodist" = "SMU",
  "Middle Tennessee State" = "Middle Tennessee",
  "Int\\'l" = "International"
  )
  )
```

Last, I'm going to set up a dataset that I can use for predicitive modeling. 
I'm also going to remove some missing values. These are due to FCS teams, 
which I do not have data for.

```{r}
dt <- regSeason %>%
  left_join(home, by = 'teamName') %>%
  left_join(opp, by = c('opponent' = 'teamName'))
dt <- na.omit(dt)
```


### Training the Models



```{r}
#splitting the dataset into a training and testing set

set.seed(82)
i <- createDataPartition(dt$score,
                        p = .75,
                        list = FALSE)
trainSet <- dt[i,-c(1:3,5)]
testSet <- dt[-i,-c(1:3,5)]
```


```{r, message=FALSE}
# Benchmark test against average score model
RMSE(testSet$home_points_per_game, testSet$score)
R2(testSet$home_points_per_game, testSet$score)

vars <- paste0(names(dt)[-c(1:5)],collapse = " + ")
vars <- sub("+ home_points_per_game ", "", vars)
formula <- as.formula(paste0("score ~ ",vars))
formula

# Linear Stepwise Model
modelLM <- caret::train(formula, trainSet, method = 'lmStepAIC',
                        trace = FALSE)

# Score Stepwise Test Set
predLM <- predict(modelLM, testSet)
RMSE(predLM,testSet$score)
R2(predLM,testSet$score)

# Gradient Boosted Trees
modelGBM <- caret::train(formula, trainSet, method = 'gbm',
                         verbose = FALSE)

# Score GBM Test Set
predGBM <- predict(modelGBM, testSet)
RMSE(predGBM,testSet$score)
R2(predGBM,testSet$score)
```

### Comparing the Predictions

```{r}
resultsPage <- 'https://www.sports-reference.com/cfb/years/2017-bowls.html'
bowlResults <- read_html(resultsPage) %>%
  html_table() %>%
  getElement(1) %>%
  setNames(c('date',
             'gametime',
             'bowl',
             'winner',
             'wpts',
             'loser',
             'lpts',
             'location',
             'na'))
bowlResults <- bowlResults[,-9]
bowlResults$winner <- str_replace_all(
  bowlResults$winner,
  c(
  "Central Florida" =  "UCF",
  "Texas Christian" = "TCU",
  "Middle Tennessee State" = "Middle Tennessee"
  )
  )

bowlResults$loser <- str_replace_all(
  bowlResults$loser,
  c(
  "Louisiana State" = "LSU",
  "Miami \\(FL\\)" = "Miami (Florida)",
  "Southern California" = "USC",
  "Alabama-Birmingham" = "UAB",
  "Southern Methodist" = "SMU"
  )
  )

bowlWinners <- bowlResults %>%
  left_join(home, by = c('winner' = 'teamName')) %>%
  left_join(opp, by = c('loser' = 'teamName'))
bowlLosers <- bowlResults %>%
  left_join(home, by = c('loser' = 'teamName')) %>%
  left_join(opp, by = c('winner' = 'teamName'))
```

```{r}
# Favored
23/nrow(bowlResults)

# Baseline
sum(bowlWinners$home_points_per_game >
  bowlLosers$home_points_per_game)/nrow(bowlResults)

# Stepwise Prediction
sum(predict(modelLM, bowlWinners) >
predict(modelLM, bowlLosers))/nrow(bowlResults)

# Prediction
sum(predict(modelGBM, bowlWinners) >
      predict(modelGBM, bowlLosers))/nrow(bowlResults)
```

